---
title: "Downloading FIA data"
author: "Hunter Stanke"
date: "2019"
description: Learn how to download FIA data and get it into R 
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{downloading}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Download data

The first step to using `rFIA` is to download subsets of the FIA Database. The easiest way to accomplish this is using `getFIA()`.  Using one line of code, you can download state subsets of the FIA Database, load data into your R environment, and optionally save those data to a local directory for future use. Alternatively, you may also choose to download subsets as .csv files from the [FIA DataMart](https://apps.fs.usda.gov/fia/datamart/datamart.html) and load into R using `readFIA()`.


```{r eval=FALSE}
library(rFIA)
# Download the state subset or Connecticut (requires an internet connection)
# Save as an object to automatically load the data into your current R session!
ct <- getFIA(states = 'CT', dir = '/path/to/save/data')

# Get multiple states worth of data (not saved since 'dir' is not specified)
northEast <- getFIA(states = c('ME', 'NH', 'VT', 'NY', 'CT', 'MA', 'RI'))
```

You can also download REFERENCE tables listed on the [FIA DataMart](https://research.fs.usda.gov/products/dataandtools/tools/fia-datamart) by specifying `states = 'REF'`:

```{r eval=FALSE}
# Get the forest type and forest type group reference tables
ref <- getFIA(states = 'ref', tables = c('FOREST_TYPE', 'FOREST_TYPE_GROUP'))
```

If you are downloading a large amount of data (e.g., the entire eastern US), you may not want to load it all into R immediately. In this case, specify `load=FALSE` in the call to `getFIA()`. This will ensure all data is download and saved to disk without maxing out your RAM. For more on 'big data' management, check out our [larger than RAM methods](https://github.com/doserjef/rFIA).

# Load data into R

If you used `getFIA` to download data in your current R session, then the database is likely automatically loaded into your current R session (unless `load=FALSE`). Theoretically, we could use `getFIA` to re-download FIA data every time you want to use, but that would be _very_ inefficient. Instead, we recommend you save downloaded FIA data using the `dir` argument in `getFIA` (automatically saves on download) or using `writeFIA` (saves any in-memory `FIA.Database`). Once data are saved on disk, you can quickly re-load them into R using `readFIA`:

```{r eval=FALSE}
## Load FIA Data from a local directory
db <- readFIA('/path/to/your/directory/')
```

# Loading multiple states

Need to load multiple state subsets of FIA data for regional analyses? No problem! Using `getFIA()`, specify mutiple state abbreviations in the `states` argument (e.g. `states = c('MI', 'IN', 'WI', 'IL')`). Alternatively, download individual states seperately and save them to the same directory. When multiple state subsets of data are loaded into R using `getFIA()` or `readFIA()`, subsets will be merged into a single `FIA.Database` object. This will allow you to use other `rFIA` functions to produce estimates for areas which straddle state boundaries!

Conveniently, you can selectively read state subsets from a directory containing multiple states worth of data using `readFIA()`. For example, lets say we previously used `getFIA()` to download FIA data for Washington, Oregon, and Idaho. We saved all this data to same directory, and hence pointing `readFIA()` to this directory will automatically load and merge all states. But what if we just want to load the data for Washington? Easy, use the `states` argument in readFIA:

```{r eval=FALSE}
# Download data for PNW states, but don't load the data yet
getFIA(states = c('WA', 'OR', 'ID'),
       dir = 'path/to/my/directory/',
       load = FALSE)

# A simple call to readFIA will load and merge all states
allStates <- readFIA(dir = 'path/to/my/directory/')

# But using the 'states' argument we can select individual states (or groups)
wa <- readFIA(dir = 'path/to/my/directory/', states = 'WA')

# Read WA and OR, but not ID
wa_or <- readFIA(dir = 'path/to/my/directory/', states = c('WA', 'OR'))
```

Given the massive size of the full FIA Database (~50 GB), users are cautioned to only download the subsets containing their region of interest.

# Loading specific tables

If you are only interested in loading/downloading a specific table from the FIA database, simply specify the names of those tables in the `tables` argument of `readFIA()` or `getFIA()` (e.g. specify `tables = c('TREE', 'PLOT')` for the TREE and PLOT tables). See the [FIA User Guide](https://research.fs.usda.gov/understory/forest-inventory-and-analysis-database-user-guide-nfi) for a complete description of the database.

By default, `getFIA()` and `readFIA()` only loads/downloads the portions of the database required to produce summaries with other `rFIA` functions (`common = TRUE`). This conserves memory on your machine and speeds download time. If you would like to load/download all available tables for a state, simple specify `common = FALSE` in the call to  `readFIA()` or `getFIA()`.

<br>

# The FIA.Database object

When FIA data is loaded into R with `readFIA()` or `getFIA()`, those data are stored in an `FIA.Database` object. An `FIA.Database` object is essentially just a list with each list element corresponding to an individual FIA table. Accordingly, users can access individual tables with `$`, `['']`, and `[['']]` operators:

```{r eval=FALSE}
# Access the TREE, PLOT, and COND tables
# Tree
db$TREE

# Plot
db['PLOT']

# Check spatial coverage of plots held in the database
plotFIA(db)
```

