---
title: "Incorporating spatial data"
author: "Hunter Stanke, Jeffrey W. Doser"
date: "2019 (last updated February 5, 2025)"
description: Learn how to estimate forest parameters within unique spatial zones
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{spatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Grouping estimates by user-defined areal units

Want to compute estimates within your own areal units (spatial polygons)? All `rFIA` estimator functions make this task fast and easy. Simply hand your spatial polygons to the `polys` argument of an estimator function, like `tpa` or `biomass`, and estimates will be grouped within those spatial zones. No need to worry about projections, *`rFIA` functions will reproject FIA data to match that of your input polygon.*

```{r}
library(rFIA)
# Most recent subset
riMR <- clipFIA(fiaRI)

# Group estimates by the areal units, and return as a dataframe
tpa_polys <- tpa(riMR, polys = countiesRI)

# Same as above, but return an sf multipolygon object (spatially enabled)
tpa_polysSF <- tpa(riMR, polys = countiesRI, returnSpatial = TRUE)
```

`polys` object must be of class `SpatialPolygons` (`sp` package), `SpatialPolygonsDataFrame` (`sp` package), or `MultiPolygon` (`sf` package). We strongly encourage users to use `sf` objects as the `sp` package is gradually being retired. 

# Returning estimates at the plot-level

Want to return estimates at the plot level and retain the spatial data associated with each FIA plot? Just specify `returnSpatial = TRUE` and `byPlot = TRUE` in any `rFIA` estimator function, and you've got it!

```{r}
# Spatial plots with biomass 
bio_pltSF <- biomass(riMR, byPlot = TRUE, returnSpatial = TRUE)

# Plot the results using default sf method
plot(bio_pltSF)

# Aboveground biomass/ acre (tons) for each plot
plot(bio_pltSF['BIO_ACRE'], pch = 19)
```

# Visualization

If you opted to return estimates as a spatial object (specify `returnSpatial = TRUE`), you can easily produce spatial choropleth maps with `plotFIA()`:

```{r}
# Plot distribution of Basal area/ acre across space
plotFIA(tpa_polysSF, y = BAA, legend.title = 'BAA (sq.ft/acre)', 
        legend.height = 0.8)

# You can even modify your call to y to adjust units!
# Convert sq.ft./acre to sq.m./acre
plotFIA(tpa_polysSF, y = BAA /10.7639104, legend.title = 'BAA (sq.m/acre)', 
        legend.height = 0.8)
```

# How to load spatial objects in R

There are many, many options available for working with spatial data in R. For loading shapefiles in R, we recommend using the function `st_read()` from `sf`.  

``` {r eval=FALSE}
# Read a shapefile into R
# dsn: path to the directory where the shapefile components are stored
# layer: file name of the shapefile components w/ no extension
myPoly <- st_read(dsn = 'path/to/my/folder/', layer = 'myShape')
```

For further reference on working with spatial data in R, we recommend these references: 

+ [Spatial Data Science with R and terra](https://rspatial.org/)
+ [Using Spatial Data with R](https://cengel.github.io/R-spatial/)
